BINF 6310 | Lab 2 | Jon Lee
---
title: "R Notebook"
output: html_notebook
---
Question 1

In a population, there is 1/3 chance that a given person has a mutation in some gene

You sample 30 people; what are the odds that exactly 12 of the people have the mutation?

```{r}
dbinom(12, 30, 1/3)
```

In R plot a probability density function (with dbinom) that shows the distribution for observing exactly (0,1,2,…30) people with the mutation.

```{r}
people <- seq(0,31,1)
dist <- dbinom(people, 30, 1/3)
plot(people, dist, type = "l", xlab = "# of people", ylab = "probability", main = "Probability for 'n' number of people having the gene")
```

What is the mean and variance for the expected number of people with the mutation.

```{r}
geneMean <- 30*(1/3)
geneMean

geneVar <- 30*(1/3)*(2/3)
geneVar
```

---
Question 2

The background expected survival rate for a disease is 0.4. You are running a clinical trial. You have 100 patients on a new drug. 47 patients die (and 53 survive)

From the bionomial test:
(2A) Plot out the probability density function with the x-axis the number of patients that survive under the null hypothesis. 

```{r}
#the null hypthesis is that the survial rate is = 0.4
people <- seq(0,100,1)
dist <- dbinom(people, 100, 0.4)
plot(people, dist, type = "l", xlab = "# of people", ylab = "probability", main = "Distribution such that the null hypothesis is true")
```

(2B) What is the p-value for a null hypothesis that the drug has no effect. (Show the one line of R code that produces this p-value)

```{r}
binom.test(53, 100, p = 0.4, alternative = "two.sided")
```

(2C) What is the p-value for a null hypothesis that the drug does not improve survival. (show you can get the same answer with binom.test(….) and sum(dbinom(….))

```{r}
binom.test(53, 100, p = 0.4, alternative = "greater")

pbinom(52, 100, 0.4, lower.tail = FALSE)
sum(dbinom(53:100, 100, 0.4))
```

---
Question 3

(3A)	Use the rbiom function to simulate 1,000 experiments in which 10,000 patients are sampled with a 1/2 chance of seeing a mutation. (You should get 1,000 numbers back with each # the # of patients from the 10,000 that had the mutation…)

(What is the one line of r-code that would produce myVals?)

```{r}
myVals <- rbinom(1000, 10000, 0.5)
myVals
```

(3B) What is the expected mean and variance of the vector in (3A).  Show that the actual mean and variance are close to the expected mean and variance.

```{r}
mean(myVals)
var(myVals)

exMean <- 10000*0.5
exMean

exVar <- 10000*0.5*0.5
exVar
```

(3C) Take the vector that results from (3A).  For each element in that vector, calculate a p-value with binom.test(….)$p.value for the null hypothesis that the frequency of the allele in the population for that experiment is 1/2.

```{r}
myPVals <- vector(length=length(myVals), mode = "double")

for(i in 1:length(myVals))
{
  myPVals[i] <- binom.test(myVals[i], 10000, alternative = "two.sided")$p.value
}

myPVals
```

Graph the histogram of all of those p-values.  
What distribution would you expect?  Is that what you see?

```{r}
hist(myPVals, xlab = "p-values", main = "Histograms of p-values")

#The resulting histrogram resembles a uniform distribution which is to be expected becuase if the null hypothesis of the experiement is true than the resulting p-values will resemble a uniform distribution
```

(3D) Change the expected value of 1/2 in (3C) to some other value. What happens to the p-values in the histogram. Would you expect the same shape of the p-value histogram with expected values of .49 as with .51?  Why or why not?

```{r}
myPValsD_0.49 <- vector(length=length(myVals), mode = "double")

for(i in 1:length(myVals))
{
  myPValsD_0.49[i] <- binom.test(myVals[i], 10000, p = 0.49, alternative = "two.sided")$p.value
}

myPValsD_0.49
```
```{r}
hist(myPValsD_0.49, xlab = "p-values", main = "Histograms of p-values with Mean = 0.49")

#We don't see a uniform distribution becuase the null hypothesis is not true, because we generated the data under a different expected value, therefor the resulting p-value are mainly small (same for 0.51)
```

```{r}
myPValsD_0.51 <- vector(length=length(myVals), mode = "double")

for(i in 1:length(myVals))
{
  myPValsD_0.51[i] <- binom.test(myVals[i], 10000, p = 0.51, alternative = "two.sided")$p.value
}

myPValsD_0.51
```

```{r}
hist(myPValsD_0.51, xlab = "p-values", main = "Histograms of p-values with Mean = 0.51")

#same as above in 0.49 expected value
```
